Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces

### 2014-01-02 papers/icwe14/icwe14.tex
minor update

\begin{abstract}

\end{abstract}

@@

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}. The last two lines shows how this API is used by end developers: by passing the \scalacode{Input} value as parameter to the function, it fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.

@@

\begin{figure}
\begin{lstlisting}[label=lst-dt-dom]
trait ElementName {
  type Element
}

def createElement(name: ElementName): name.Element = ...
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-events]
trait EventName {
  type Data
}
object Click extends EventDef { type Data = MouseEvent }

def on(name: EventName)(handler: Rep[name.Data] => Rep[Unit]): Rep[Unit]

val clicks = on(Click)
clicks(event => println(event.offsetX))
\end{lstlisting}
\end{figure}

@@

@@

>>>>

\begin{abstract}
Encoding dynamically typed APIs in statically typed languages can be challenging. We present new ways to encode them in order to preserve type safety and to keep the same expression power.
\end{abstract}

@@

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}. The last two lines shows how this API is used by end developers: by passing the \scalacode{Input} value as parameter to the function, it fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.

@@

@@

Our encoding makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript, but with type safety.

However, every function taking an element name or an event name as parameter must have type parameters too, making its type signature harder to read.

This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}. Essentially, the idea is that, instead of using a type parameter, we use a type member. Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala. The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.

\begin{figure}
\begin{lstlisting}[label=lst-dt-dom]
trait ElementName {
  type Element
}
object Div extends ElementName { type Element = DivElement }
object Input extends ElementName { type Element = InputElement }

def createElement(name: ElementName): name.Element
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-events]
trait EventName {
  type Event
}
object Click extends EventDef { type Event = MouseEvent }

trait EventTarget {
  def addEventListener(name: EventName)(handler: name.Event => Unit): Unit
}
\end{lstlisting}
\end{figure}

Implementing listings \ref{lst-js-comb} and \ref{lst-js-react} using this encoding is straightforward, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.

\begin{figure}
\begin{lstlisting}[label=lst-dt-comb]
def createAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
  val element = document.createElement(tagName)
  element.addEventListener(eventName)(event => listener(event, element))
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-react]
def observe(target: EventTarget, name: EventName) =
  (listener: (name.Event => Unit)) => target.addEventListener(name)(listener)
\end{lstlisting}
\end{figure}

With this encoding, the functions using event names or element names are not anymore cluttered with type parameters.

\subsection{Implementation in js-scala}

We implemented our pattern in js-scala~\cite{Kossakowski12_JsDESL}, a Scala library providing JavaScript code generators\footnote{Source code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.

@@

We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using \emph{generics} or path-dependent types. Our encoding gives more type safety than existing solutions, while keeping the same expression power and modularity level as the native API.
